A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-30 paper.tex
More text moving
Tweaks to simplification section

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary or phylogenetic tree is a necessary prerequisite for many analyses of
species~\citep{rannala2003genetics}, non-recombining sections of
DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
and viruses~\citep{grenfell2004science}.
A phylogenetic tree captures information about the genealogy
of the sample in an elegant and concise
way, by postulating a set of ancestors common to the samples
together with paths of inheritance between them. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}.
However, the genetic genealogy of a recombining organism
cannot be described as a single evolutionary tree: instead,
the inheritance of each local section of the genome may be described
by a (slightly) different tree~\citet{hudson1983properties}. Inferring complex
genealogies of this nature
presents profound technical difficulties, meaning tree-based approaches are
rarely used in population and statistical genetics. Therefore until recently, analysis of
population genetic data has primarily focussed on summaries derived
from the local trees~\citep{tajima1983evolutionary,tavare1984line}
rather than using the underlying genealogy itself. Such
derived measures include allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
principal components~\citep{mcvean2009genealogical}.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical,hejase2022deep}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which risks hampering
progress: we currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies. This leads to basic errors in statements
about what different inference methods produce as output, as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

\section*{Ancestral graphs: a brief history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
It is a stochastic process, where each random realisation
is a genealogical tree describing the genetic ancestry of the samples.
Numerous elaborations have been incorporated into the
basic process~\citep{hudson1990gene,hein2004gene,wakely2008coalescent};
in particular, early work by \citet{hudson1983properties}
analysed the effects of recombination on the random genealogical
trees generated by the coalescent. This
generated several fundamental analytical results
as well as describing the elegant algorithm
at the heart of simulation
programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}
used in thousands of studies.
In its original formulation, Hudson's algorithm produced a
sequence of genealogical trees along the genome.
Critically, although the correlation structure among these
local trees was analysed analytically, those correlations
were not captured explicitly: the ancestry
was viewed as a series of disconnected local trees rather than
as a single cohesive object.

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
in which the stochastic process itself is encoded as a graph,
and showed how the local trees generated by Hudson's algorithm
were embedded within
it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to this graph,
representing both the stochastic process and
its random realisations, as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to describe recombinant ancestry as single object that
could be reasoned about and inferred in a way
that was missing from Hudson's treatments.

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
It is important to note that such recombination-parsimony ARGs
have quite different properties from realisations of the
ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
Firstly, branch lengths are not typically estimated,
% why do we need the following two lines
and node times are required in order to compute a likelihood under the
coalescent with recombination (see section XXX).
Secondly, even if branch lengths were estimated,
the realisations would have a very small likelihood since
the coalescent with recombination is often highly \emph{un}parsimonious
in terms of recombination events.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was therefore broken.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation},
although not universal, with authors also using the term
in the original stochastic
process sense~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent}.

% Gusfield~\citep{gusfield2014recombinatorics}
% defines also defines an ARG as a concrete realisation
% and studies instances that satisfy particular
% properties such as ``galled trees''
% \citep{wang2001perfect, gusfield2004optimal}
% as well as describing algorithms for more general phylogenetic
% networks~\citep{huson2010phylogenetic}.

In parallel with this tendency to decouple the term
Ancestral Recombination Graph from the stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes described as a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode desired
patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
% Could mention the Ancestral Gene Transfer graph here too, within
% something like "Similarly the Ancestral Gene Transfer Graph [cite]
% and [other ancestral graphs, if they exist?] model genealogical
% processes in bacteria as a graph.

The point of this historical overview is to illustrate that the term
Ancestral Recombination Graph has different meanings in different
contexts, and in its original form it was closely coupled to
the coalescent with recombination. In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent marginal trees on the right.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{k} has three children, and the genomes
\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
children.
The penultimate local tree is highlighted to aid discussion.
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
}
\end{figure}

\section*{Simplification}\label{ARG_simplification}
ARG simplification~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy, while maintaining the crucial properties
that the ``meaningful'' topology in the local trees is not affected,
and the relationship between nodes in adjacent trees is maintained.
The idea of meaningful local tree topologies is based on the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative nodes from the local trees by modifying the gARG.

@@

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
edge annotations.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \textsf{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node,
something which is impossible to encode in an eARG.

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \textsf{k} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, \emph{all}
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, it removes potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.

@@

>>>>

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary or phylogenetic tree is a necessary prerequisite for many analyses of
species~\citep{rannala2003genetics}, non-recombining sections of
DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
and viruses~\citep{grenfell2004science}.
A phylogenetic tree captures information about the genealogy
of the sample in an elegant and concise
way, by postulating a set of ancestors common to the samples
together with paths of inheritance between them. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}.
However, the genetic genealogy of a recombining organism
cannot be captured by a single tree: instead,
the inheritance of each section of the genome may be described
by a (slightly) different local tree~\citep{hudson1983properties}. Inferring complex
genealogies of this nature
presents profound technical difficulties, meaning genealogy-based approaches have
not been widely used in population and statistical genetics. Until recently, analysis of
population genetic data has therefore primarily focussed on summaries derived
from the local trees~\citep{tajima1983evolutionary,tavare1984line}
rather than using the full underlying genealogy itself. Such
derived measures include allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
principal components~\citep{mcvean2009genealogical}.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogy-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as inputs have begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical,hejase2022deep}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which risks hampering
progress: we currently lack a well-defined data-model and shared terminology
to discuss genealogies in the presence of recombination. This leads to basic errors in statements
about what different inference methods produce as output, as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

\section*{Ancestral graphs: a brief history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
It is a stochastic process, where each random realisation
is a genealogical tree describing the genetic ancestry of the sample.
Numerous extensions to the model have been proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}, incorporating various types of evolutionary processes. 
In particular, \citet{hudson1983properties}
analysed the effects of recombination on coalescent trees, generating several fundamental analytical results,
as well as describing the elegant algorithm
still at the heart of widely-used simulation
programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}.
In its original formulation, Hudson's algorithm produced a
sequence of genealogical trees along the genome.
Crucially, although the correlation structure among these
local trees was analysed analytically, those correlations
were not captured explicitly: the ancestry
was viewed as a set of disconnected local trees rather than
as a single cohesive object.

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
formulating it as a stochastic process where each realisation 
is a genealogy encoded as a graph,
and showed that the local trees generated by Hudson's algorithm
are embedded within this graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to both the stochastic process and
its random realisations as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to reason about and infer recombinant ancestry as a single object,
which was not possible within Hudson's framework.

Subsequent work on ARGs proceeded in broadly three main directions: focussing on (1) 
exploring the mathematical properties of the coalescent with recombination and 
related stochastic processes, (2) inferring evolutionary parameters under 
(approximations to) this model, either with or without explicitly reconstructing the 
genealogy of the sample, and (3) treating the ARG as a discrete graph, ignoring the 
generating stochastic process, and studying its properties from a computational and 
algorithmic perspective.

% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
An extensive body of work has developed in
studying the coalescent with recombination 
% Can we add a couple of the most well known CwR examples here
and other related 
graph-valued stochastic processes from a mathematical perspective.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection instead of recombination.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode
desired patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
% Could mention the Ancestral Gene Transfer graph here too, within
% something like "Similarly the Ancestral Gene Transfer Graph [cite]
% and [other ancestral graphs, if they exist?] model genealogical
% processes in bacteria as a graph.

% Inference of parameters and explicit ARG reconstruction
Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
Several breakthroughs in this direction were achieved through
formulating simplified but more tractable approximations to the full 
model~\citep{mcvean2005approximating,marjoram2006fast,li2011inference,
paul2011accurate,schiffels2014inferring}.
The related problem of \emph{sampling} genealogies compatible with a given 
dataset under the coalescent with recombination also proved notoriously difficult
computationally; progress in explicitly inferring genealogies at scale 
has similarly been achieved through resorting to principled
approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
or moving away from the coalescent with recombination altogether and seeking
to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.

% Optimisation problems to do with ARGs and other phylogenetic networks
There has also been substantial interest in formulating and answering 
fundamental questions about properties
of the ARG as a discrete graph structure, abstracting away from the
generating process and focussing on the ARG topology without considering 
branch lengths. 
The first prominent problem was calculating (lower bounds on) the minimum number of
recombinations required to reconstruct a valid genealogy for a given
sample~\citep{myers2003bounds}, and constructing the corresponding 
minimal (parsimonious)
ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
focus has been on algorithmic and combinatorial results, that largely do not
have direct relevance to the inference problems described above.

%It is important to note that parsimonious ARGs
%have quite different properties from realisations of the
%ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
%Firstly, branch lengths are not typically estimated,
%% why do we need the following two lines
%and node times are required in order to compute a likelihood under the
%coalescent with recombination (see section XXX).
%Secondly, even if branch lengths were estimated,
%the realisations would have a very small likelihood since
%the coalescent with recombination is often highly \emph{un}parsimonious
%in terms of recombination events.

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the 
stochastic process generating genealogies in the presence of 
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent}, 
as well as, more commonly, the concrete realisations of the 
ancestry~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent local trees on the right.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{k} has three children, and the genomes
\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple
children.
The penultimate local tree is highlighted to aid discussion.
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
}
\end{figure}

\section*{Simplification}\label{ARG_simplification}
ARG simplification~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy,
retaining the ``meaningful'' topology within local trees
and maintaining the identity of nodes across different trees.
The idea of meaningful local tree topologies is based on the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right.
For example, consider the path above node
\textsf{a} in the highlighted tree (covering positions 60 to 70)
in Fig.~\ref{fig-simplification}A.
Before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes.
This reflects the
passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For most purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative nodes from the local trees by modifying the gARG.

@@

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
edge annotations.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \textsf{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \textsf{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node.
% something which is impossible to encode in an eARG.

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \textsf{k} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.

@@

% JK TODO: finish up this section with a quick version of the
% para here. Basically say, people often say an ARG is a
% set of local trees cite cite, but it's not as simple as that as we
% can see from the example (probably).
